---
title: "Likelihood estimation of a normal-multinomial distribution"
author: "Marc Comas-Cufí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Likelihood estimation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library(normalmultinomial)
set.seed(1)
```

# The normal-multinomial distribution

The normal-multinomial distribution is the distribution resulting after compounding the normal distribution defined on the Simplex and the well-known multinomial distribution. The parameters of a normal-multinomial distribution are the mean and covariance, $\mu$ and $\Sigma$, which define the parameters of the underlying normal distribution.

For example, if we consider the parameters

```{r}
mu = c(1, 0)
Sigma = matrix(c(1, 0,
                 0, 1), nrow=2)
```

a possible sample of 10 observations is given by

```{r, warning=F}
rnormalmultinomial(mu = mu, sigma = Sigma, size = rep(20, 10))
```


### The probability density function (pdf) 

The pdf of a normal-multinomial distribution for $k+1$ components is

\[
f(x; \mu, \Sigma) = \int_{a \in \mathbb{R}^{k}} MVN(a ;\mu, \Sigma) \frac{(x_1+\dots+x_{k+1})!}{x_1! \dots x_{k+1}!} (\frac{1}{\omega(a)})^{x_{k+1}}\prod_{\ell=1}^k (\frac{e^{a_\ell}}{\omega(a)})^{x_{\ell}} da
\]

where $a = (a_1, \dots, a_k)$ and $\omega(a) = 1 + e^{a_1} + \dots + e^{e_k}$. Equivalently in a more compact notation,

\[
f(x; \mu, \Sigma) = \frac{(x_1+\dots+x_{k+1})!}{x_1! \dots x_{k+1}!} \int_{a \in \mathbb{R}^{k}} MVN(a ;\mu, \Sigma)  \omega(a)^{-x_1 - \dots - x_{k+1}} e^{x_1 a_1 + \dots + x_k a_k} da.
\]

### Maximum likelihood estimation

Estimating the parameters of a normal-multinomial distribution from the pdf by maximum likelihood seems not straightforward. The pdf for the join distribution of $(X,A)$ is

\[
f(x, a; \mu, \Sigma) = \frac{(x_1+\dots+x_{k+1})!}{x_1! \dots x_{k+1}!} MVN(a ;\mu, \Sigma) \omega(a)^{-x_1 - \dots - x_{k+1}} e^{x_1 a_1 + \dots + x_k a_k},
\]

which seems more tractable than $f(x; \mu, \Sigma)$. Therefore, if we knew the value of the observations $a_1, \dots, a_n$, estimating the parameters $\mu$ and $\Sigma$ from a sample $x_1, \dots, x_n$ will be straighforward. In fact, given $a_1, \dots, a_n$, estimating $\mu$ and $\Sigma$ is direct by
\[
\hat{\mu} = \bar{a}
\]

and 

\[
\hat{\Sigma} = Cov(a).
\]

### EM-algorithm

To estimate the parameters $\mu$ and $\sigma$ a reasonable approach is to consider an EM algorithm to obtain $(X,A)$ where $X$ is the observed sample and $A$ is not observed.


### E-step

Suppose $\mu$ and $\Sigma$ are known, let  $\mu^{(s)}$ and $\Sigma^{(s)}$ be the current estimations. We need to calculate

\[
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ \sum_{i=1}^n log \left( f(x_i, a_i; \mu, \Sigma) \right) \right].
\]

First, we are going to separate $log \left( f(x, a; \mu, \Sigma) \right)$ in four parts

\[
\sum_{i=1}^n log \left( f(x_i, a_i; \mu, \Sigma) \right) = M_1(X) + M_2(A, \mu, \Sigma) + M_3(A, X) + M_4(A, X)
\]

where

\[
\begin{align}
M_1(X) &= \sum_{i=1}^n log \left( \frac{(x_{i,1}+\dots+x_{i,k+1})!}{x_{i,1}! \dots x_{i,k+1}!} \right) \\
M_2(A, \mu, \Sigma) &= \sum_{i=1}^n log \left(  MVN(a_i ;\mu, \Sigma) \right) \\
M_3(A, X) &= \sum_{i=1}^n log \left( \omega(a_i)^{-x_{i,1} - \dots - x_{i,k+1}} \right) \\
M_4(A, X) &= \sum_{i=1}^n x_{i,1} a_{i,1} + \dots + x_{i,k} a_{i,k}
\end{align}
\]

The expected value of each part is given by

\[
\begin{align}
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_1(X) \right] &= M_1(X), \\
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_2(A, \mu, \Sigma) \right] &=  E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ -\frac{n}{2}|\Sigma| -\frac{1}{2} tr(\Sigma^{-1} \sum_{i=1}^n a_i a_i^T)  -\frac{n}{2} \mu' \Sigma^{-1}\mu +  \mu' \Sigma^{-1} \sum_{i=1}^n a_i   \right] = \\
&= -\frac{n}{2}|\Sigma| -\frac{1}{2} tr(\Sigma^{-1} \sum_{i=1}^n E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[a_i a_i^T \right])  -\frac{n}{2} \mu' \Sigma^{-1}\mu +  \mu' \Sigma^{-1} \sum_{i=1}^n E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[a_i\right] \\
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_3(A, X) \right] &= \sum_{i=1}^n -(x_{i,1} + \dots + x_{i,k+1}) E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ log \omega(a_i) \right] \\
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_4(A, X) \right] &= \sum_{i=1}^n (x_{i,1}, \dots, x_{i,k})^T E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right]
\end{align}
\]


Therefore, to compute $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ \sum_{i=1}^n log \left( f(x_i, a_i; \mu, \Sigma) \right) \right]$ we need to calculate 

* $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right]$, 
* $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[a_i a_i^T \right]$ and 
* $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ log \omega(a_i) \right]$.

### M-step

Once $Q(\mu, \Sigma) := E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_1(X) + M_2(A, \mu, \Sigma) + M_3(A, X) + M_4(A, X) \right]$ has been calculated in the E-step, we should find the parameters $\mu^{(s+1)}$, $\Sigma^{(s+1)}$ which maximize $Q(\mu, \Sigma)$. At this point, note that to optimize $Q(\mu, \Sigma)$ we only need to optimize expression $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_2(A, \mu, \Sigma) \right]$, the other terms are constant with respect to $\mu$ and $\Sigma$. Therefore, in the E-step it is not necessary to evaluate $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ log \omega(a_i) \right]$.


With the expected values $m_{i,1} = E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right]$ and $m_{i,2} = E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i a_i^T \right]$ calculated in the E-step, we get the estimators 

\[
\begin{align}
\mu^{(s+1)} &= \frac{\sum_{i=1}^n m_{i,1}}{n}, \\
\Sigma^{(s+1)} &=  \sum_{i=1}^n \frac{m_{i,2}}{n} - \mu^{(s+1)} {\mu^{(s+1)}}^T
\end{align}
\]

which maximize $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ M_2(A, \mu, \Sigma) \right]$.

#### $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right]$ estimation

To calculate $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right]$ we need to calculate

\[
\begin{align}
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i \right] &= \int_{a_i \in \mathbb{R}^k} a_i f(a_i, x_i; \mu^{(s)}, \Sigma^{(s)}) da_i \\
&= \int_{a_i \in \mathbb{R}^k} a_i  \frac{(x_{i,1}+\dots+x_{i, k+1})!}{x_{i,1}! \dots x_{i, k+1}!} MVN(a_i ;\mu^{(s)}, \Sigma^{(s)}) \omega(a_i)^{-x_{i,1} - \dots - x_{i, k+1}} e^{x_{i,1} a_{i,1} + \dots + x_{i,k} a_{ik}} d a_i \\
&= \frac{(x_{i,1}+\dots+x_{i, k+1})!}{x_{i,1}! \dots x_{i, k+1}!} \int_{a_i \in \mathbb{R}^k} a_i \omega(a_i)^{-x_{i,1} - \dots - x_{i, k+1}}   e^{x_{i,1} a_{i,1} + \dots + x_{i,k} a_{ik}} d MVN(a_i ;\mu^{(s)}, \Sigma^{(s)}),
\end{align}
\]

#### $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i a_i^T \right]$ estimation

To calculate $E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i a_i^T \right]$ we need to calculate

\[
\begin{align}
E_{A/(\mu^{(s)}, \Sigma^{(s)})}\left[ a_i a_i^T \right] &= \int_{a_i \in \mathbb{R}^k} a_i a_i^T f(a_i, x_i; \mu^{(s)}, \Sigma^{(s)}) da_i \\
&= \int_{a_i \in \mathbb{R}^k} a_i a_i^T  \frac{(x_{i,1}+\dots+x_{i, k+1})!}{x_{i,1}! \dots x_{i, k+1}!} MVN(a_i ;\mu^{(s)}, \Sigma^{(s)}) \omega(a_i)^{-x_{i,1} - \dots - x_{i, k+1}} e^{x_{i,1} a_{i,1} + \dots + x_{i,k} a_{ik}} d a_i\\
&= \frac{(x_{i,1}+\dots+x_{i, k+1})!}{x_{i,1}! \dots x_{i, k+1}!} \int_{a_i \in \mathbb{R}^k} a_i a_i^T \omega(a_i)^{-x_{i,1} - \dots - x_{i, k+1}} e^{x_{i,1} a_{i,1} + \dots + x_{i,k} a_{ik}} dMVN(a_i ;\mu^{(s)}, \Sigma^{(s)}),
\end{align}
\]