---
title: "Generating Normal-Multinomial observations"
author: "Marc Comas-Cufí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(1)
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, results = 'hold', collapse = T, comment = "#>")
```

This vignette explains how to generate Normal-Multinomial observations using the package `normalmultinomial`.

```{r}
library(normalmultinomial)
```

## Normal and Multinomial generators

You can generate normal and multinomial observations using the functions `rnormal` 

```{r, results='hold'}
MU = c(0,0)
SIGMA = matrix(c(2,1,1,2), nrow=2)
(A <- rnormal(n = 10, mu = MU, sigma = SIGMA))
```

and `rmultinomial`

```{r, results='hold'}
(X.multinomial <- rmultinomial(n = 10, size = c(5, 50), prob = c(1,1,1)))
```

## Normal-Multinomial generator

Combining functions `rnormal` and `rmultinomial` function you can generate Normal-Multinomial observations

```{r}
(X.normalmultinomial <- rmultinomial(n = 10, size = c(5, 50), prob = exp(cbind(A, 0))))
```

Using the function `rnormalmultinomial` function, you can generate Normal-Multinomial observations directly

```{r}
(X.normalmultinomial <- rnormalmultinomial(n = 10, size = c(5, 50), mu = MU, sigma = SIGMA))
```
